---
title: "Figure 3B - PLSDA of two-month infant antibody profiles"
output: html_document
---

This notebook applies repeated LASSO feature selection and PLSDA to compare
the five maternal vaccination-timing groups at the two-month infant visit. Run
`01_load_data.Rmd` first.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(ropls)
library(systemsseRology)

helper_dir <- if (dir.exists("../helpful_functions")) {
  "../helpful_functions"
} else {
  "helpful_functions"
}
source(file.path(helper_dir, "data_io.R"))
source(file.path(helper_dir, "analysis_functions.R"))
source(file.path(helper_dir, "plotting_functions.R"))

stopifnot(all(vapply(
  c("infant_data_M2", "infant_meta_M2", "infant_persist_M2"),
  exists,
  logical(1)
)))
set.seed(2023)
```

## 1. Select and scale serology features

Features detected in more than 30% of two-month samples are retained.

```{r prepare-features}
cov_antigens <- paste(
  c(
    "RBD_beta", "RBD_delta", "RBD_gamma", "RBD_omicron", "S1", "S2",
    "Spike_alpha", "Spike_beta", "Spike_delta", "Spike_gamma",
    "Spike_omicron", "RBD_alpha", "RBD_WT", "Spike_WT"
  ),
  collapse = "|"
)

serology_data <- select_serology_features(
  infant_data_M2,
  include = cov_antigens
)
serology_data <- filter_detected_features(
  serology_data,
  infant_persist_M2,
  minimum_fraction = 0.30
)

X <- scale(serology_data, center = TRUE, scale = TRUE)
y <- droplevels(infant_meta_M2$unique_groups)
```

## 2. Configure feature annotations and colors

```{r plot-options}
group_colors <- c(
  "Group 1" = "#669ccb",
  "Group 2" = "#004488",
  "Group 3" = "#eccc64",
  "Group 4" = "#92414e",
  "Group 5" = "#e69aa3"
)

feature_info <- make_plsda_feature_metadata(
  colnames(X),
  colors = list(group = group_colors)
)

plot_options <- list(
  df_features = feature_info$metadata,
  loading_alpha = 1,
  score_alpha = 1,
  LV_ind = c(1, 2),
  color_features = "antigen",
  colors = feature_info$colors,
  y_name = "group",
  alpha = 1,
  size = 2.5,
  stroke = 0.5
)
```

## 3. Fit the full-feature PCA and PLSDA models

```{r full-models}
pca_model <- pca_ropls(X)
pca_scores <- visualize_ropls_scores(pca_model, y, options = plot_options)

plsda_options <- list(n_LV = 2)
full_plsda_model <- train_ropls(X, y, options = plsda_options)
full_plsda_scores <- visualize_ropls_scores(
  full_plsda_model,
  y,
  options = plot_options
) +
  theme(axis.title = element_text(size = 13, face = "bold"))
```

## 4. Repeated LASSO feature selection and final PLSDA

LASSO selection is repeated 100 times. Features selected in at least 80% of
trials are used in the final two-latent-variable PLSDA model.

```{r selected-model}
selection_options <- list(
  n_trials = 100,
  threshold = 0.80,
  return_count = FALSE
)
selected_features <- select_repeat(
  X,
  y,
  selector = select_lasso,
  options = selection_options
)

X_selected <- X[, selected_features, drop = FALSE]
selected_model <- train_ropls(X_selected, y, options = plsda_options)

selected_scores <- visualize_ropls_scores(
  selected_model,
  y,
  options = plot_options
) +
  theme(axis.title = element_text(size = 18, face = "bold"))

selected_loadings <- visualize_ropls_loadings(
  selected_model,
  options = plot_options
) +
  theme(axis.title = element_text(size = 18, face = "bold"))

loading_options <- plot_options
loading_options$X <- X_selected
loading_options$y <- y
loading_options$LV_ind <- 1
loading_options$mark_enrichment <- TRUE
selected_loading_bars <- visualize_ropls_loadings_bar(
  selected_model,
  options = loading_options
) +
  theme(
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 12, face = "bold")
  )
```

## 5. Summarize variable importance

```{r vip}
vip_table <- make_vip_table(selected_model, X, y, minimum_vip = 1)

vip_plot <- ggplot(vip_table, aes(label, VIP, fill = enriched)) +
  geom_col(color = "black") +
  scale_fill_manual(values = group_colors) +
  labs(x = NULL, y = "VIP score") +
  theme_classic() +
  theme(
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none"
  )
```

## 6. Validate the selected-feature model

Validation uses 10-fold cross-validation, 100 label permutations, and 10
repeated validation trials.

```{r model-validation}
validation_selection_options <- list(
  n_trials = 100,
  threshold = 0.30,
  return_count = FALSE,
  alpha = 1
)
lasso_selector <- function(X, y) {
  select_repeat(
    X,
    y,
    selector = select_lasso,
    options = validation_selection_options
  )
}

validation_method <- list(
  select = lasso_selector,
  train = train_ropls,
  predict = predict_ropls,
  score = score_accuracy
)
validation_options <- list(
  n_folds = 10,
  rf_trials = 0,
  pt_trials = 100,
  save = TRUE,
  compare_pred = "y"
)

validation_results <- validate_repeat(
  X,
  y,
  validation_method,
  validation_options,
  n_trials = 10
)
validation_plot <- visualize_validate(
  validation_results,
  options = list(y_label = "accuracy")
) +
  theme(
    axis.title.y = element_text(size = 18),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 18)
  )
```

## 7. Save model figures and selected features

```{r save-results}
ggsave(output_path("Figure_3", "Figure_3B_PLSDA_scores.png"), selected_scores)
ggsave(output_path("Figure_3", "Figure_3B_PLSDA_loadings.png"), selected_loadings)
ggsave(output_path("Figure_3", "Figure_3B_VIP.png"), vip_plot)
ggsave(output_path("Figure_3", "Figure_3B_validation.png"), validation_plot)

write.csv(
  data.frame(feature = selected_features),
  output_path("Figure_3", "Figure_3B_selected_features.csv"),
  row.names = FALSE
)
```
