---
title: "Supplementary Figure 3 - UMAP of two-month infant profiles"
output: html_document
---

This notebook performs an unsupervised UMAP analysis of two-month infant
antibody profiles and examines associations with vaccination group, infant
sex, and selected antibody features. Run `Figure_3/01_load_data.Rmd` first.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggprism)
library(ggpubr)
library(rstatix)
library(Seurat)

helper_dir <- if (dir.exists("../helpful_functions")) {
  "../helpful_functions"
} else {
  "helpful_functions"
}
source(file.path(helper_dir, "data_io.R"))
source(file.path(helper_dir, "analysis_functions.R"))

stopifnot(all(vapply(
  c("infant_data_M2", "infant_meta_M2", "infant_persist_M2"),
  exists,
  logical(1)
)))
set.seed(2023)
```

## 1. Select antibody features

```{r prepare-features}
cov_antigens <- paste(
  c(
    "RBD_beta", "RBD_delta", "RBD_gamma", "RBD_omicron", "S1", "S2",
    "Spike_alpha", "Spike_beta", "Spike_delta", "Spike_gamma",
    "Spike_omicron", "RBD_alpha", "RBD_WT", "Spike_WT"
  ),
  collapse = "|"
)
umap_data <- select_serology_features(infant_data_M2, include = cov_antigens)
umap_data <- filter_detected_features(
  umap_data,
  infant_persist_M2,
  minimum_fraction = 0.30
)

rownames(umap_data) <- infant_meta_M2$Pedi_label
umap_metadata <- infant_meta_M2
rownames(umap_metadata) <- infant_meta_M2$Pedi_label
```

## 2. Run PCA, clustering, and UMAP

```{r run-umap}
seurat_object <- CreateSeuratObject(counts = t(umap_data))
seurat_object@meta.data <- umap_metadata
seurat_object <- FindVariableFeatures(seurat_object, selection.method = "vst")
seurat_object <- ScaleData(seurat_object, features = rownames(seurat_object))
seurat_object <- RunPCA(
  seurat_object,
  features = VariableFeatures(seurat_object),
  npcs = 40
)
seurat_object <- FindNeighbors(seurat_object, dims = 1:20)
seurat_object <- FindClusters(seurat_object, resolution = 0.5)
seurat_object <- RunUMAP(
  seurat_object,
  dims = 1:20,
  n.neighbors = 30,
  min.dist = 0.2
)

cluster_markers <- FindAllMarkers(
  seurat_object,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
```

## 3. Assemble UMAP coordinates and metadata

```{r umap-data-frame}
umap_coordinates <- Embeddings(seurat_object[["umap"]])
umap_plot_data <- cbind(seurat_object[[]], umap_coordinates)
umap_feature_data <- cbind(umap_data, umap_plot_data)

group_colors <- c(
  "Group 1" = "#669ccb",
  "Group 2" = "#004488",
  "Group 3" = "#eccc64",
  "Group 4" = "#92414e",
  "Group 5" = "#e69aa3"
)
sex_colors <- c("M" = "#00798C", "F" = "#d1495b")
cluster_levels <- levels(umap_plot_data$seurat_clusters)
cluster_colors <- grDevices::colorRampPalette(c("#edae49", "#66a182"))(
  length(cluster_levels)
)
names(cluster_colors) <- cluster_levels

umap_theme <- theme_classic() +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.55, face = "bold", size = 18)
  )
```

## 4. Plot sample annotations

```{r annotation-plots}
umap_by_group <- ggplot(
  umap_plot_data,
  aes(UMAP_1, UMAP_2, fill = unique_groups)
) +
  geom_point(shape = 21, size = 5) +
  scale_fill_manual(values = group_colors) +
  labs(title = "Infant 2 months", fill = NULL) +
  umap_theme

umap_by_sex <- ggplot(
  umap_plot_data,
  aes(UMAP_1, UMAP_2, fill = Fetal_Sex)
) +
  geom_point(shape = 21, size = 5) +
  scale_fill_manual(values = sex_colors) +
  labs(title = "Infant 2 months", fill = "Infant sex") +
  umap_theme

umap_by_cluster <- ggplot(
  umap_plot_data,
  aes(UMAP_1, UMAP_2, fill = seurat_clusters)
) +
  geom_point(shape = 21, size = 5) +
  scale_fill_manual(values = cluster_colors) +
  labs(title = "Infant 2 months", fill = "Cluster") +
  umap_theme
```

## 5. Plot selected antibody features

```{r feature-plots}
features_to_plot <- c("IgG1_Spike_WT", "IgG2_Spike_WT")
umap_feature_plots <- setNames(lapply(features_to_plot, function(feature) {
  ggplot(
    umap_feature_data,
    aes(UMAP_1, UMAP_2, fill = .data[[feature]])
  ) +
    geom_point(shape = 21, size = 5) +
    scale_fill_gradientn(
      colors = colorspace::diverge_hcl(5),
      name = "log10 (MFI)"
    ) +
    labs(title = format_feature_label(feature)) +
    umap_theme
}), features_to_plot)

umap_feature_panel <- ggarrange(
  plotlist = umap_feature_plots,
  nrow = 2,
  ncol = 1
)
```

## 6. Compare selected features by infant sex

```{r sex-comparisons}
sex_comparison_data <- cbind(infant_data_M2, infant_meta_M2)

sex_plots <- setNames(lapply(features_to_plot, function(feature) {
  plot_data <- data.frame(
    value = sex_comparison_data[[feature]],
    Fetal_Sex = sex_comparison_data$Fetal_Sex
  )
  test <- wilcox_test(plot_data, value ~ Fetal_Sex) %>%
    add_significance("p") %>%
    add_xy_position(x = "Fetal_Sex")

  ggplot(plot_data, aes(Fetal_Sex, value, fill = Fetal_Sex)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
    add_pvalue(test, label = "p.signif", tip.length = c(0, 0)) +
    scale_fill_manual(values = c("F" = "#F88379", "M" = "#189ad3")) +
    labs(
      x = "Infant sex",
      y = "log10 (MFI)",
      title = format_feature_label(feature)
    ) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}), features_to_plot)

sex_comparison_panel <- ggarrange(plotlist = sex_plots, legend = "none", nrow = 2)
```

## 7. Save supplementary panels

```{r save-results}
ggsave(output_path("Sup_Fig_3", "UMAP_vaccine_group.png"), umap_by_group)
ggsave(output_path("Sup_Fig_3", "UMAP_infant_sex.png"), umap_by_sex)
ggsave(output_path("Sup_Fig_3", "UMAP_clusters.png"), umap_by_cluster)
ggsave(output_path("Sup_Fig_3", "UMAP_features.png"), umap_feature_panel)
ggsave(output_path("Sup_Fig_3", "sex_comparisons.png"), sex_comparison_panel)
write.csv(
  cluster_markers,
  output_path("Sup_Fig_3", "UMAP_cluster_markers.csv"),
  row.names = FALSE
)
```
