---
title: "Supplementary Figure 4 - PLSDA of six-month infant profiles"
output: html_document
---

This notebook compares six-month infant antibody profiles across the three
maternal vaccination-timing groups. Run `Figure_4/01_load_data.Rmd` first.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggsci)
library(RColorBrewer)
library(ropls)
library(systemsseRology)

helper_dir <- if (dir.exists("../helpful_functions")) {
  "../helpful_functions"
} else {
  "helpful_functions"
}
source(file.path(helper_dir, "data_io.R"))
source(file.path(helper_dir, "analysis_functions.R"))
source(file.path(helper_dir, "plotting_functions.R"))

stopifnot(all(vapply(
  c("temp_data", "temp_meta", "sel_feat", "infant_persist_M6"),
  exists,
  logical(1)
)))
set.seed(2023)
```

## 1. Prepare the three vaccination groups

```{r prepare-groups}
group_levels <- c("<20 weeks", "20 - 28 weeks", ">28 weeks")
keep <- temp_meta$Vax_group %in% group_levels
analysis_meta <- temp_meta[keep, , drop = FALSE]
analysis_data <- temp_data[keep, , drop = FALSE]
y <- factor(analysis_meta$Vax_group, levels = group_levels)

simpsons_colors <- ggsci::pal_simpsons()(5)
group_colors <- c(simpsons_colors[2], simpsons_colors[1], simpsons_colors[5])
names(group_colors) <- group_levels
```

## 2. Filter and scale serology features

Features detected in at least 30% of the selected samples are retained.

```{r prepare-features}
serology_data <- analysis_data[, sel_feat, drop = FALSE]
detection_data <- infant_persist_M6[keep, , drop = FALSE]
serology_data <- filter_detected_features(
  serology_data,
  detection_data,
  minimum_fraction = 0.30
)

X <- scale(serology_data, center = TRUE, scale = TRUE)
```

## 3. Configure feature annotations and colors

```{r plot-options}
feature_info <- make_plsda_feature_metadata(
  colnames(X),
  colors = list(`Vax1 Group` = group_colors)
)

plot_options <- list(
  df_features = feature_info$metadata,
  loading_alpha = 1,
  score_alpha = 1,
  LV_ind = c(1, 2),
  color_features = "antigen",
  colors = feature_info$colors,
  y_name = "Vax1 Group",
  alpha = 1,
  size = 2.5,
  stroke = 0.5
)
```

## 4. Fit full and selected-feature PLSDA models

LASSO selection is repeated 100 times. Features selected in at least 30% of
trials are used in the final model.

```{r fit-models}
plsda_options <- list(n_LV = 2)
full_model <- train_ropls(X, y, options = plsda_options)
full_scores <- visualize_ropls_scores(full_model, y, options = plot_options)

selection_options <- list(
  n_trials = 100,
  threshold = 0.30,
  return_count = FALSE
)
selected_features <- select_repeat(
  X,
  y,
  selector = select_lasso,
  options = selection_options
)

X_selected <- X[, selected_features, drop = FALSE]
selected_model <- train_ropls(X_selected, y, options = plsda_options)

selected_scores <- visualize_ropls_scores(
  selected_model,
  y,
  options = plot_options
) +
  theme(axis.title = element_text(size = 18, face = "bold"))

selected_loadings <- visualize_ropls_loadings(
  selected_model,
  options = plot_options
) +
  theme(axis.title = element_text(size = 18, face = "bold"))

loading_options <- plot_options
loading_options$X <- X_selected
loading_options$y <- y
loading_options$LV_ind <- 1
loading_options$mark_enrichment <- TRUE
selected_loading_bars <- visualize_ropls_loadings_bar(
  selected_model,
  options = loading_options
)
```

## 5. Summarize variable importance

```{r vip}
vip_table <- make_vip_table(selected_model, X, y, minimum_vip = 1)

vip_plot <- ggplot(vip_table, aes(label, VIP, fill = enriched)) +
  geom_col(color = "black") +
  scale_fill_manual(values = group_colors) +
  labs(x = NULL, y = "VIP score") +
  theme_classic() +
  theme(
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none"
  )
```

## 6. Validate the model

```{r model-validation}
lasso_selector <- function(X, y) {
  select_repeat(
    X,
    y,
    selector = select_lasso,
    options = selection_options
  )
}
validation_method <- list(
  select = lasso_selector,
  train = train_ropls,
  predict = predict_ropls,
  score = score_accuracy
)
validation_options <- list(
  n_folds = 10,
  rf_trials = 0,
  pt_trials = 100,
  save = TRUE,
  compare_pred = "y"
)

validation_results <- validate_repeat(
  X,
  y,
  validation_method,
  validation_options,
  n_trials = 10
)
validation_plot <- visualize_validate(
  validation_results,
  options = list(y_label = "accuracy")
)
```

## 7. Save supplementary results

```{r save-results}
ggsave(output_path("Sup_Fig_4", "PLSDA_scores.png"), selected_scores)
ggsave(output_path("Sup_Fig_4", "PLSDA_loadings.png"), selected_loadings)
ggsave(output_path("Sup_Fig_4", "PLSDA_VIP.png"), vip_plot)
ggsave(output_path("Sup_Fig_4", "PLSDA_validation.png"), validation_plot)

write.csv(
  data.frame(feature = selected_features),
  output_path("Sup_Fig_4", "PLSDA_selected_features.csv"),
  row.names = FALSE
)
```
