---
title: "02 - Prepare analysis objects"
author: "Paola Lopez"
output: html_document
---

This notebook applies the manuscript cohort rules, derives vaccination-timing
groups, and saves the maternal, cord-blood, and infant objects used by the
figure notebooks.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)

helper_dir <- if (dir.exists("../helpful_functions")) {
  "../helpful_functions"
} else {
  "helpful_functions"
}
source(file.path(helper_dir, "data_io.R"))
source(file.path(helper_dir, "data_preprocessing.R"))

intermediate_file <- input_path("preprocessing_intermediate.RData")
assert_files_exist(intermediate_file)
load(intermediate_file)
```

## 1. Apply cohort exclusions

Maternal vaccination with or without prior COVID-19 infection is retained.
Infant samples collected after infant infection are excluded. Samples whose
second maternal vaccine dose occurred after delivery are also excluded.

```{r cohort-selection}
keep_infant <- new_infant_meta$Sample_collected_after_baby_COVID != "Yes"
infant_meta <- new_infant_meta[keep_infant, , drop = FALSE]
infant_data <- new_infant_data[keep_infant, , drop = FALSE]

mom_meta <- new_mom_meta
mom_data <- comb_mom_data
CB_meta_v2 <- cord_blood_meta
CB_data_v2 <- comb_CB_data
```

## 2. Derive COVID-19 vaccination-timing groups

```{r covid-vaccine-groups}
add_covid_groups <- function(metadata) {
  metadata$Vax1_group <- classify_covid_vaccination(
    metadata$GA_Vax1_days,
    metadata$GA_delivery_days
  )
  metadata$Vax2_group <- classify_covid_vaccination(
    metadata$GA_Vax2_days,
    metadata$GA_delivery_days
  )
  metadata
}

infant_meta <- add_covid_groups(infant_meta)
mom_meta <- add_covid_groups(mom_meta)

infant_match <- match(CB_meta_v2$PediID, new_infant_meta$PediID)
CB_meta_v2$GA_Vax1_days <- new_infant_meta$GA_Vax1_days[infant_match]
CB_meta_v2$GA_Vax2_days <- new_infant_meta$GA_Vax2_days[infant_match]
CB_meta_v2$GA_delivery_days <- new_infant_meta$GA_delivery_days[infant_match]
CB_meta_v2 <- add_covid_groups(CB_meta_v2)

keep_infant <- infant_meta$Vax2_group != "Post-preg"
infant_data <- infant_data[keep_infant, , drop = FALSE]
infant_meta <- infant_meta[keep_infant, , drop = FALSE]

keep_maternal <- mom_meta$Vax2_group != "Post-preg"
mom_data <- mom_data[keep_maternal, , drop = FALSE]
mom_meta <- mom_meta[keep_maternal, , drop = FALSE]

keep_cord <- CB_meta_v2$Vax2_group != "Post-preg"
CB_data_v2 <- CB_data_v2[keep_cord, , drop = FALSE]
CB_meta_v2 <- CB_meta_v2[keep_cord, , drop = FALSE]
```

## 3. Derive influenza and Tdap timing groups

```{r other-vaccine-groups}
add_other_vaccine_groups <- function(metadata) {
  metadata$Flu_vax_group <- classify_pregnancy_vaccination(
    metadata$GA_flu_days,
    metadata$GA_delivery_days
  )
  metadata$TDAP_vax_group <- classify_pregnancy_vaccination(
    metadata$GA_TDAP_days,
    metadata$GA_delivery_days
  )
  metadata$Flu_vax_group <- collapse_outside_pregnancy(metadata$Flu_vax_group)
  metadata$TDAP_vax_group <- collapse_outside_pregnancy(metadata$TDAP_vax_group)
  metadata
}

infant_meta <- add_other_vaccine_groups(infant_meta)
mom_meta <- add_other_vaccine_groups(mom_meta)

CB_meta_v2$GA_flu_days <- new_infant_meta$GA_flu_days[infant_match][keep_cord]
CB_meta_v2$GA_TDAP_days <- new_infant_meta$GA_TDAP_days[infant_match][keep_cord]
CB_meta_v2 <- add_other_vaccine_groups(CB_meta_v2)

flu_levels <- c(
  "Before Pregnancy",
  "Before 20 weeks",
  "Between 20 to 28 weeks",
  "After 28 weeks",
  "Outside Pregnancy"
)

infant_meta$Flu_vax_group <- factor(infant_meta$Flu_vax_group, levels = flu_levels)
mom_meta$Flu_vax_group <- factor(mom_meta$Flu_vax_group, levels = flu_levels)
CB_meta_v2$Flu_vax_group <- factor(CB_meta_v2$Flu_vax_group, levels = flu_levels)

infant_meta$TDAP_vax_group <- factor(
  infant_meta$TDAP_vax_group,
  levels = c("Before 20 weeks", "Between 20 to 28 weeks", "After 28 weeks")
)
mom_meta$TDAP_vax_group <- factor(
  mom_meta$TDAP_vax_group,
  levels = c("Between 20 to 28 weeks", "After 28 weeks")
)
CB_meta_v2$TDAP_vax_group <- factor(
  CB_meta_v2$TDAP_vax_group,
  levels = c("Between 20 to 28 weeks", "After 28 weeks")
)
```

## 4. Label the five joint vaccination groups

```{r joint-vaccine-groups}
infant_meta$unique_groups <- label_covid_vaccine_groups(
  infant_meta$Vax1_group,
  infant_meta$Vax2_group
)
mom_meta$unique_groups <- label_covid_vaccine_groups(
  mom_meta$Vax1_group,
  mom_meta$Vax2_group
)
CB_meta_v2$unique_groups <- label_covid_vaccine_groups(
  CB_meta_v2$Vax1_group,
  CB_meta_v2$Vax2_group
)
```

## 5. Validate and save the analysis inputs

```{r save-analysis-objects}
stopifnot(
  nrow(infant_data) == nrow(infant_meta),
  nrow(mom_data) == nrow(mom_meta),
  nrow(CB_data_v2) == nrow(CB_meta_v2)
)

save_rdata("infant_data", "infant_data.RData")
save_rdata("infant_meta", "infant_metadata.RData")
save_rdata("mom_data", "maternal_data.RData")
save_rdata("mom_meta", "maternal_metadata.RData")
save_rdata("CB_data_v2", "cord_data.RData")
save_rdata("CB_meta_v2", "cord_metadata.RData")
save_rdata("neg_control_threshold", "negative_control_thresholds.RData")
```
