---
title: "Figure 1 - Maternal and infant antibody persistence"
output: html_document
---

This notebook compares wild-type Spike and RBD antibody measurements across
maternal blood, cord blood, and infant samples collected at two and six months.
Run `01_load_data.Rmd` first.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggprism)
library(ggpubr)
library(ggsci)
library(rstatix)
library(tidyr)

helper_dir <- if (dir.exists("../helpful_functions")) {
  "../helpful_functions"
} else {
  "helpful_functions"
}
source(file.path(helper_dir, "data_io.R"))
source(file.path(helper_dir, "analysis_functions.R"))
source(file.path(helper_dir, "plotting_functions.R"))

required_objects <- c(
  "mom_data", "mom_meta", "CB_data_v2", "CB_meta_v2",
  "infant_data_M2", "infant_meta_M2", "infant_data_M6", "infant_meta_M6",
  "neg_control_threshold"
)
stopifnot(all(vapply(required_objects, exists, logical(1))))
```

## 1. Combine the four sample types

```{r combine-samples}
sample_blocks <- list(
  make_sample_block(mom_data, mom_meta, "mom", "Infant_gender"),
  make_sample_block(CB_data_v2, CB_meta_v2, "cord", "Infant_gender"),
  make_sample_block(infant_data_M2, infant_meta_M2, "infant", "Fetal_Sex"),
  make_sample_block(
    infant_data_M6,
    infant_meta_M6,
    "infant 6 months",
    "Fetal_Sex"
  )
)
merged_samples <- do.call(rbind, sample_blocks)

sample_levels <- c("mom", "cord", "infant", "infant 6 months")
feature_data <- select_serology_features(
  merged_samples,
  include = "Spike_WT|RBD_WT"
)
long_data <- make_feature_long(merged_samples, feature_data, sample_levels)
```

## 2. Test differences and calculate detection counts

Kruskal-Wallis tests are performed feature-by-feature. The resulting p-values
are adjusted across features using the Benjamini-Hochberg FDR procedure.

```{r statistics}
feature_tests <- kruskal_by_feature(long_data)
long_data <- add_detection_status(long_data, neg_control_threshold)
detection_summary <- summarize_detection(long_data)
```

## 3. Plot longitudinal measurements

```{r plot-features}
igg_features <- c(
  paste0(c("IgG", "IgG1", "IgG2", "IgG3", "IgG4"), "_Spike_WT"),
  paste0(c("IgG", "IgG1", "IgG2", "IgG3", "IgG4"), "_RBD_WT")
)
fc_features <- c(
  paste0(c("FcRn", "FcgR2AR", "FcgR2B", "FcgR3AV", "FcgR3B"), "_Spike_WT"),
  paste0(c("FcRn", "FcgR2AR", "FcgR2B", "FcgR3AV", "FcgR3B"), "_RBD_WT")
)

make_x_labels <- function(feature) {
  counts <- detection_summary[detection_summary$feature == feature, ]
  c(
    mom = paste0("M\n", counts$mom),
    cord = paste0("C\n", counts$cord),
    infant = paste0("2 m.\n", counts$infant),
    `infant 6 months` = paste0("6 m.\n", counts$`infant 6 months`)
  )
}

annotation_start <- function(feature) {
  if (grepl("IgG2|IgG3", feature)) 5.8 else if (grepl("IgG4", feature)) 6.0 else 6.7
}

type_colors <- ggsci::pal_npg()(4)
all_features <- c(igg_features, fc_features)
feature_plots <- setNames(lapply(all_features, function(feature) {
  plot_longitudinal_feature(
    long_data = long_data,
    feature = feature,
    thresholds = neg_control_threshold,
    colors = type_colors,
    x_labels = make_x_labels(feature),
    y_positions = seq(annotation_start(feature), by = 0.55, length.out = 10)
  )
}), all_features)

figure_1_igg <- ggarrange(
  plotlist = feature_plots[igg_features],
  nrow = 2,
  ncol = 5
)
figure_1_fc <- ggarrange(
  plotlist = feature_plots[fc_features],
  nrow = 2,
  ncol = 5
)
```

## 4. Save figure panels

```{r save-figures}
ggsave(
  output_path("Figure_1", "Figure_1A_IgG.png"),
  figure_1_igg,
  width = 10,
  height = 3.6,
  units = "in"
)
ggsave(
  output_path("Figure_1", "Figure_1A_Fc_receptors.png"),
  figure_1_fc,
  width = 10,
  height = 3.6,
  units = "in"
)
```
